Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-03-12 paper.tex

Genetic models have been hampered in their contribution to this discussion
because they primarily assume (or, at least, have been tested under) a
tree-like model of isolation-with-migration. Alternative theoretical scenarios
have been proposed, such as stepping stone models \citep{Arredondo2021-qa} or
population coalescence and fragmentation \citep{Scerri2019-xg}, but these
approaches are more challenging to interpret and fit to data.
However, new population genetic tools now allow for inference involving tens to
hundreds of genomes from multiple populations and greater complexity
\citep{Kamm2020-vn,Ragsdale2019-nt,Speidel2019-nj}. Inspired by evidence for
Neanderthal admixture with humans in Eurasia, several recent articles
have shown that introducing an archaic hominin ghost population contributing to African
populations in the period surrounding the Out-of-Africa migration event
substantially improves the description of genetic data relative to
single-origin models
\citep{Plagnol2006-lt,Hammer2011-bx,Hsieh2016-gk,Hey2018-pw,Ragsdale2019-nt,Durvasula2020-td,Lorente-Galdos2019-vz,Durvasula2020-td}.
This has driven speculation about the geographic range of this ghost
population, possible links to specific fossils, and the possibility of
finding ancient DNA evidence \citep{Hsieh2016-gk}. However, these prior
articles share two weaknesses. First, they only contrast a single-origin model
with an archaic hominin admixture model, leaving out other plausible models
\citep{Henn2018-rf} (Figure~\ref{fig:proposed-models}).
Second, they
focus on a small subset of African diversity, either because of small sample
sizes (2-5 genomes) or because they rely on 1000 Genomes data which only
recruited populations of recent West African or Bantu-speaking ancestry
(Figure~\ref{fig:diversity}C). 
While ancient DNA from Eurasia has helped us understand early human history
outside of Africa, there is no comparably ancient DNA to elucidate early history
in Africa \citep{Lipson2022-xf}.

>>>>

Genetic models have been hampered in their contribution to this discussion
because they primarily assume (or, at least, have been tested under) a
tree-like model of isolation-with-migration. Alternative theoretical scenarios
have been proposed, such as stepping stone models \citep{Arredondo2021-qa} or
population coalescence and fragmentation \citep{Scerri2019-xg}, but these
approaches are more challenging to interpret and fit to data.
However, new population genetic tools now allow for inference involving tens to
hundreds of genomes from multiple populations and greater complexity
\citep{Kamm2020-vn,Ragsdale2019-nt,Speidel2019-nj}. Inspired by evidence for
Neanderthal admixture with humans in Eurasia, several recent articles
have shown that introducing an archaic hominin ghost population contributing to African
populations in the period surrounding the Out-of-Africa migration event
substantially improves the description of genetic data relative to
single-origin models
\citep{Plagnol2006-lt,Hammer2011-bx,Hsieh2016-gk,Hey2018-pw,Ragsdale2019-nt,Lorente-Galdos2019-vz,Durvasula2020-td}.
This has driven speculation about the geographic range of this ghost
population, possible links to specific fossils, and the possibility of
finding ancient DNA evidence (e.g.\citep{Hsieh2016-gk}). However, these prior
articles share two weaknesses. First, they only contrast a single-origin model
with an archaic hominin admixture model, leaving out other plausible models
\citep{Henn2018-rf} (Figure~\ref{fig:proposed-models}).
Second, they
focus on a small subset of African diversity, either because of small sample
sizes (2-5 genomes) or because they rely on 1000 Genomes data which only
recruited populations of recent West African or Bantu-speaking ancestry
(Figure~\ref{fig:diversity}C). 
While ancient DNA from Eurasia has helped us understand early human history
outside of Africa, there is no comparably ancient DNA to elucidate early history
in Africa \citep{Lipson2022-xf}.

### 2022-03-12 supplement.tex

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-omni-hapmap-comparison.pdf}
    \caption{
        \textbf{Comparison of LD statistics parsed using two recombination maps}
        The recombination map determines genetic distances between pairs of SNPs,
        so using different maps could result in systematic biases in parsed
        statistics if the maps differ significantly. Here, the solid lines are
        statistics parsed using the HapMapII recombination map, and dashed lines
        are using the OMNI-YRI map. Shading represents 95\% confidence intervals
        from the OMNI map. These two maps results in LD statistics that are
        indistinguishable from one another. Notations for each statistic are
        described in section~\ref{sec:supp-ld-stats} and indexed by populations
        (0: Nama, 1: Mende, 2: Gumuz, 3: Amhara/Oromo, 4: British).
    }
    \label{fig:supp-omni-hapmap-comparison}
\end{figure}

>>>>

\begin{figure}[ht]
    \centering
    \includegraphics{figures/mergers.pdf}
    \caption{
        \textbf{Two scenarios for population rearrangement.}
        In our demographic models, we considered
        A) continuous migration models, in which one of the stem population expands 
        (splits into contemporary populations), and the other stem population(s)
        has continuous symmetric migration with those populations;
        and B) one or more of the stem populations expands, with instantaneous pulse
        (or ``merger'') events from the other stem population,
        so that recent populations are formed by mergers
        of multiple ancestral populations.
    }
    \label{fig:mergers}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-omni-hapmap-comparison.pdf}
    \caption{
        \textbf{Comparison of LD statistics parsed using two recombination maps.}
        The recombination map determines genetic distances between pairs of SNPs,
        so using different maps could result in systematic biases in parsed
        statistics if the maps differ significantly. Here, the solid lines are
        statistics parsed using the HapMapII recombination map, and dashed lines
        are using the OMNI-YRI map. Shading represents 95\% confidence intervals
        from the OMNI map. These two maps results in LD statistics that are
        indistinguishable from one another. Notations for each statistic are
        described in section~\ref{sec:supp-ld-stats} and indexed by populations
        (0: Nama, 1: Mende, 2: Gumuz, 3: Amhara/Oromo, 4: British).
    }
    \label{fig:supp-omni-hapmap-comparison}
\end{figure}
